Spit-Apply-Combine with Collapsing Groups

### 2022-11-23 paper/paper.tex
typos

Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
$U$ and returns a single value in some domain $X$. Here, $U$ is our data set
and $\phi$ an aggregating function. We split $U$ into groups using the
following notation. Let $A$ be finite set that has no more elements than $U$,
and let $f$ be a function that takes an element of $U$ and returns a value in
$A$. We can think of $A$ as a set of group labels, and $f$ as the function that
assigns a label to each element of $U$. This way, $f$ divides $U$ into
non-overlapping groups, since every element in $U$ can only get one label (that
is just how mathematical functions work). In formal terminology we say that
$f:U\to A$ is a \emph{partition} of $U$.

@@

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set pairs $(a,x)\in
A\times X$ }

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
Here, the output is collected in a set $R$. We also introduced some new
notation: the notation $2^U$ stands for `the set of all subsets of $U$'. The
notation $f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you
give it a set of labels from $A$ and it returns all the elements of $U$ that
have any of those labels. This definition is slightly too general for the
current algorithm, but it will come in handy when we move to collapsing groups.

Let us see how the elements $U$, $f\to A$ and $\phi$ are implemented
in practice. Consider the signature of the R \texttt{aggregate} function for data
frames (we skip arguments that are not important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \texttt{x} is a data frame where each column is a variable to be
aggregated.  The argument \texttt{by} represents the label(s) for each record
that are to be used for grouping. So the function $f\to A$ is implemented by
forcing the user to make sure that the position of each label in \texttt{by}
corresponds to the row number in the data frame \texttt{x}. The argument
\texttt{FUN} is the function that aggregates each subset of \texttt{x}.  Here is
an example of how a user might call this function from the R prompt.
%
\begin{verbatim}
> aggregate(iris["Sepal.Length"], by=iris["Species"], FUN=mean)
     Species Sepal.Length
1     setosa        5.006
2 versicolor        5.936
3  virginica        6.588
\end{verbatim}
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The output
also reveals in the first column the set $A$: each row corresponds to a unique
value in \texttt{Species} column.

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each 
subset of a dataset, possibly using values from a larger
subset. The input of the algorithm consists again of a
finite set $U$, and aggregation function $\phi$ that takes
a subset of $U$ and returns a value in some domain $X$.
Compared to Algorithm~\ref{alg:sac} we need two extra
inputs. First, we need a user-defined function that checks whether
a given subset $d$ of $U$ is suitable for computing $\phi(d)$.
We will call this function $\beta: 2^U\to \mathbb{B}$, where
$\mathbb{B}=\{\texttt{True},\texttt{False}\}$. 
Typical tests are checking whether there are sufficient records
available, or whether certain variables have a low enough fraction
of missing values. Second, we need
a \emph{collapsing scheme}
\begin{equation}
U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{center}
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\end{center}
Here, $A_0$ is the most detailed partition, splitting a dataset into seven
groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
$0113$ and the second, labeled $01.2$ is constructed by combining
$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.

@@

To generalize this algorithm to generic collapsing schemes, we introduce
the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
returns the corresponding label in $A_k$. This can formally be defined 
as consecutive application of $f_1, f_2,\cdots, f_k$ to an element of 
$A_0$. In notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to that label via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally
write down our main Algorithm~\ref{alg:saccg}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The data frame interface}

>>>>

Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
$U$ and returns a value in some domain $X$. Here, $U$ represents a data set and
$\phi$ an aggregating function. We split $U$ into groups using the following
notation. Let $A$ be finite set that has no more elements than $U$, and let $f$
be a function that takes an element of $U$ and returns a value in $A$. We can
think of $A$ as a set of group labels, and $f$ as the function that assigns a
label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
groups, since every element in $U$ can only get one label.  In formal
terminology we say that $f:U\to A$ is a \emph{partition} of $U$.

@@

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs $(a,x)\in
A\times X$ }

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
Here, the output is collected in a set $R$ (this set is technically a function
$A\to X$ since it pairs for each $a\in A$ with one element of $X$). We also
introduced some new notation: the notation $2^U$ stands for `the set of all
subsets of $U$'. The notation $f^{-1}$ stands for the formal \emph{inverse} of
$f$. That is: you give it a set of labels from $A$ and it returns all the
elements of $U$ that have any of those labels. This definition is slightly too
general for the current algorithm, but it will come in handy when we move to
collapsing groups.

Let us see how the elements $U$, $f:U\to A$, and $\phi$ are implemented
in practice. Consider the signature of the R \texttt{aggregate} function for data
frames (we skip arguments that are not important for the discussion).
\begin{verbatim}
  aggregate(x, by, FUN)
\end{verbatim}
Here, \texttt{x} is a data frame where each column is a variable $U$ to be
aggregated.  The argument \texttt{by} represents the record-wise label(s) that
are used for grouping. So the function $f\to A$ is implemented by forcing the
user to make sure that the position of each label in \texttt{by} corresponds to
the row number in the data frame \texttt{x}. The argument \texttt{FUN} is the
function $\phi$ that aggregates each subset of \texttt{x}.  Here is an example of how
a user might call this function from the R prompt.
%
\begin{verbatim}
> aggregate(iris["Sepal.Length"], by=iris["Species"], FUN=mean)
     Species Sepal.Length
1     setosa        5.006
2 versicolor        5.936
3  virginica        6.588
\end{verbatim}
%
Note that the correspondence in position of the \texttt{Species} label and the
record position is implemented by taking them from the same data frame. The output
also reveals in the first column the set $A$: each row corresponds to a unique
value in \texttt{Species} column.

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each part of a dataset,
possibly using values from a larger subset. The input of the algorithm consists
again of a finite set $U$ and an aggregation function $\phi$ that takes a subset
of $U$ and returns a value in some domain $X$.  Compared to
Algorithm~\ref{alg:sac} we need two extra inputs. First, we need a user-defined
function that checks whether a given subset $d$ of $U$ is suitable for
computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
checking whether there are sufficient records available, or whether certain
variables have a low enough fraction of missing values. Second, we need a
\emph{collapsing scheme}
\begin{equation}
U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{equation}
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\label{eq:tablerep}
\end{equation}
Here, $A_0$ is the most detailed partition, splitting a dataset into seven
groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
$0113$ and the second, labeled $01.2$ is constructed by combining
$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.

@@

In order to generalize this algorithm towards generic collapsing schemes, we
introduce the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and
returns the corresponding label in $A_k$. This can formally be defined as
consecutive application of $f_1, f_2,\cdots, f_k$ to an element of $A_0$. In
notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally write
down our main Algorithm~\ref{alg:saccg}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The data frame interface}
\eqref{eq:tablerep}
